\thispagestyle{empty}

\begin{center}
    \textbf{\huge{Acknowledgements}}
\end{center}
\vspace{1cm}

The first person I would like to thank is my supervisor, Dr. Sanjib Dey, for his undeterred support and guidance throughout this work. Our discussions on physics, as well as life played a crucial role in seeing this dissertation to fruition. I am indebted to his invaluable inputs and critical comments of my work. \par 

I express my heartfelt gratitude to Dr. Yogesh Singh, and Dr. Abhishek Chaudhuri for pushing me to complete this work even when I doubted myself and was effectively lost. Their constant nudges were essential in keeping me motivated. \par 

My life at IISER Mohali for the past year would not have been complete without my friends who made me laugh even in the darkest of times. For them I would like revise Professor Albus Dumbledore's words from Harry Potter and the Deathly Hallows to say, ``Help would always be given at IISER Mohali, to those who deserve and cherish it. Thank you Vidhika, Sourav, Sandip, Sougata, Rajeshree, Abhimanyu, Antriksh, Ayan, Nilaj, Abhinandan, Bhavya, Sanat, Babai, Subhashree Di, Gaurav, Deepraj, Yuvraj, Rahul, Monu, Nikhil for all the cherishable memories. \par

I would not have come this far without the immense sacrifice and patience of my Maa, and Papa. Your believe in me and my passion for science allowed me to keep going even in times of despair. Thank you for all the unconditional love, and privileges that you have offered me and keep giving.   